Theory of phonon polaritons in polar dielectrics

### 2018-08-02 paper.tex
some paper edits on 2d and nonlocal

Before applying this formalism to a real material and also before discussion of the two-dimensional case, we briefly comment on "nonlocality" in the phonon contribution to the dielectric function. Note that Equation (12) has a wavevector dependence which is implicit through the phonon dispersion. As a result, the phonon contribution to the dielectric function sees contributions from the entire phonon band structure, including the large wavevector components. However, $\epsilon(\mathbf{q},\omega)$ measures the response to an electromagnetic perturbation of frequency $\omega$ and wavevector $\mathbf{q}$. In the vast majority of experiments that have been performed on phonon polaritons, the wavevector used to probe the polariton is on the order of $\frac{\omega}{c}$ which is much less than the wavevector scale at which the optical phonon dispersion varies, which is of the order of $\frac{\pi}{a}$ where $a$ is a lattice constant of the polar dielectric.  Even in scattering near-field optical microscope experiments, where one uses an AFM tip of radius of about $20$ nm to probe the polaritons, the wavevectors accessed would still be extremely small compared to the extent of the Brillouin zone. It would then seem almost hopeless to see nonlocal behavior for the phonon-polaritons. However, a recent experiment \cite{iranzo2018probing} has shown that plasmons in graphene, by means of a gold mirror, can be confined to dimensions of about 1 nm in the dimension transverse to the graphene sheet. Additionally, another recent experiment \cite{benz2016single} has leveraged so-called "picocavities", sub-nanometer gaps between a metal nanoparticle and metal film, to access very strong light-matter couplings and extreme variations in fields. These two experiments show a way to get large optically-accessed wavevectors to probe nonlocal behavior in polar dielectrics.

@@

\section{Linear response of phonon polaritons in two dimensions}
We now move to generalize the derivation of the previous section to describe the optical phonon contribution to Maxwell's equations coming from a two-dimensional material. Because of the change of symmetry in the system, we will not be able to perform the analysis by Fourier transforming Maxwell's equations in three dimensions. However, we will apply a technique which is often used to analyze electromagnetic modes in 2D systems such as plasmons in 2D conductors. An example of the technique is shown in detail in \cite{jablan2009plasmonics}, but we summarize the essential details here.

@@

To connect this relation to the discussion of the three-dimensional case, we write a microscopic form for the polarization response function of the ionic lattice. It is simply
\begin{equation}
\boldsymbol{\Pi}(\mathbf{q},\omega) =  \frac{1}{\epsilon_0 A}\sum\limits_{m,n}\frac{\mathbf{P}_{mn}(\mathbf{q})\otimes\mathbf{P}_{nm}(\mathbf{q})}{\hbar\omega + E_{nm}+i0^+}\left(e^{\beta E_m}-e^{\beta E_n} \right),
\end{equation}
where the only change is to replace a normalization volume with a normalization area $A$. Here, $\mathbf{P}(\mathbf{q})$ is the 2D Fourier transform of the polarization density. As the form of the displacement of the ions is still given by Equation (3), $\mathbf{P}(\mathbf{q})$ is still given by Equations (7-10), but with the Born charges and eigendisplacements appropriate to the 2D system of interest. Note that although the lattice is two dimensional, in general, the ions may be displaced in any three directions and thus $\boldsymbol{\Pi}$ is still a $3\times3$ matrix. Following the same kind of reasoning that lead to Equation (13), if we parameterize the effect of the different masses and Born charges in the unit cell by $M^*$ and $Q^*$, we will find a form for the polarization response function which is given by the second term of Equation (13) but with $V$ replaced by $A$. Defining $N/A \equiv n_s$, Equation (14) for the dispersion relation of 2D phonon polaritons tells us that
\begin{equation}
\omega = \sqrt{\omega_{TO}^2+\frac{n_sQ^{*2}}{2M^*\epsilon_0\epsilon}q}.
\end{equation}
Note that this dispersion has a strong resemblance to the plasmon dispersion relation of two-dimensional electron gases, which is normally of the form $\omega = \sqrt{\frac{n_s e^2  }{2m\epsilon_0\epsilon}q}$ \cite{stern1967polarizability}. The only difference is the replacement of electron parameters (density, charge, mass) by the corresponding lattice parameters in the phononic case, and a shift in the minimum frequency by the transverse optical phonon frequency.

\textcolor{blue}{Still looking for more to say about this. This probably requires some brainstorming discussion between us. For now, what I'll do is prepare a plot of the Fourier transform of the green's function of the Maxwell equations for the 2D sheet (p-polarized reflectivity, as was done in Argentene).}

>>>>

Before applying this formalism to a real material and also before discussion of the two-dimensional case, we briefly comment on "nonlocality" in the phonon contribution to the dielectric function. Note that Equation (12) has a wavevector dependence which is implicit through the phonon dispersion.  Therefore, it is possible that the phonon polariton resonances red-shift or blue-shift, depending on how the relative position of the phonons at finite-wavevector versus zero wavevector. For example, in 4H-SiC, upon moving along the $\Gamma-L$ direction of the Brillouin zone, the optical phonon which determines the phonon polariton, calculated to be at 764 cm$^{-1}$ at $\Gamma$, red-shifts by 2  cm$^{-1}$ at 1/10 of the way along the $\Gamma-L$ direction, by 5 cm$^{-1}$ at 1/10 of the way along the $\Gamma-L$ direction, and 13 cm$^{-1}$ at 1/2 of the way along the $\Gamma-L$ direction. This will lead to a red-shift of the Reststrahlen bands by these amounts.

%$\epsilon(\mathbf{q},\omega)$ measures the response to an electromagnetic perturbation of frequency $\omega$ and wavevector $\mathbf{q}$.  I
In the vast majority of experiments that have been performed on phonon polaritons, the wavevector used to probe the polariton is on the order of $\frac{\omega}{c}$ which is much less than the wavevector scale at which the optical phonon dispersion varies, which is of the order of $\frac{\pi}{a}$ where $a$ is a lattice constant of the polar dielectric.  Even in scattering near-field optical microscope experiments, where one uses an AFM tip of radius of about $20$ nm to probe the polaritons, the wavevectors accessed would still be extremely small compared to the extent of the Brillouin zone. It would then seem almost hopeless to see nonlocal behavior for the phonon-polaritons. However, a recent experiment \cite{iranzo2018probing} has shown that plasmons in graphene, by means of a gold mirror, can be confined to dimensions of about 1 nm in the dimension transverse to the graphene sheet. Additionally, another recent experiment \cite{benz2016single} has leveraged so-called "picocavities", sub-nanometer gaps between a metal nanoparticle and metal film, to access very strong light-matter couplings and extreme variations in fields. These two experiments show a way to get large optically-accessed wavevectors to probe nonlocal behavior in polar dielectrics.

%This is exemplified in Figure 3, where we show the ionic contribution to the permittivity (i.e. we do not consider the electronic part). We consider wavevector directions along $\Gamma-L$ and consider wavevectors that are $1/10$ and $1/4$ of the way along the Brillouin zone. We see that the major effect is to red-shift the phonon resonance by a few inverse centimeters. However, the effect "downstream", in the region where phonon polaritons are, is much smaller, with the permittivity being almost the same as in the local case.  In making this plot, we make the assumption that the variation of the Born effective charges is negligible upon moving to finite-wavevector optical phonons.

%However, $\epsilon(\mathbf{q},\omega)$ measures the response to an electromagnetic perturbation of frequency $\omega$ and wavevector $\mathbf{q}$.

@@

\section{Linear response of phonon polaritons in two dimensions}
We now move to generalize the derivation of the previous section to describe the optical phonon contribution to Maxwell's equations coming from a two-dimensional material. Unlike the previous section, we will not demonstrate the results of the calculation on a particular material. Here, we are only interested in finding an expression for the optical response of the surface in terms of ab-initio-derivable parameters, as well as the general form of the dispersion relation of phonon-polaritons in 2D.  Because of the change of symmetry in the system, we will not be able to perform the analysis by Fourier transforming Maxwell's equations in three dimensions. However, we will apply a technique which is often used to analyze electromagnetic modes in 2D systems such as plasmons in 2D conductors. An example of the technique is shown in detail in \cite{jablan2009plasmonics}, but we summarize the essential details here.

@@

To connect this relation to the discussion of the three-dimensional case, we write a microscopic form for the polarization response function of the ionic lattice. It is simply
\begin{equation}
\boldsymbol{\Pi}(\mathbf{q},\omega) =  \frac{1}{\epsilon_0 A}\sum\limits_{m,n}\frac{\mathbf{P}_{mn}(\mathbf{q})\otimes\mathbf{P}_{nm}(\mathbf{q})}{\hbar\omega + E_{nm}+i0^+}\left(e^{\beta E_m}-e^{\beta E_n} \right),
\end{equation}
where the only change is to replace a normalization volume with a normalization area $A$. Here, $\mathbf{P}(\mathbf{q})$ is the 2D Fourier transform of the polarization density. As the form of the displacement of the ions is still given by Equation (3), $\mathbf{P}(\mathbf{q})$ is still given by Equations (7-10), but with the Born charges and eigendisplacements appropriate to the 2D system of interest. Note that although the lattice is two dimensional, in general, the ions may be displaced in any three directions and thus $\boldsymbol{\Pi}$ is still a $3\times3$ matrix. Following the same kind of reasoning that lead to Equation (13), if we parameterize the effect of the different masses and Born charges in the unit cell by $M^*$ and $Q^*$, we will find a form for the polarization response function which is given by the second term of Equation (13) but with $V$ replaced by $A$. Defining $N/A \equiv n_s$, Equation (14) for the dispersion relation of 2D phonon polaritons tells us that
\begin{equation}
\omega = \sqrt{\omega_{TO}^2+\frac{n_sQ^{*2}}{2M^*\epsilon_0\epsilon}q}.
\end{equation}
Note that this dispersion has a strong resemblance to the plasmon dispersion relation of two-dimensional electron gases, which is normally of the form\cite{stern1967polarizability} $\omega = \sqrt{\frac{n_s e^2  }{2m\epsilon_0\epsilon}q}$ . The only difference is the replacement of electron parameters (density, charge, mass) by the corresponding lattice parameters in the phononic case, and a shift in the minimum frequency by the transverse optical phonon frequency.

We note that the dispersion relation of Equation (8) can in fact be obtained by taking the infinitely thin-limit of a bulk polar dielectric, and finding the dispersion of the resulting phonon polaritons in the quasi-electrostatic limit. In particular, for an isotropic polar dielectric of thickness $2d$ with boundaries at $z=\pm d$, solving Laplace's equations yields as the dispersion of the even parity mode:
\begin{equation}
\tanh(qd) = -\frac{\epsilon}{\epsilon(\omega)} \rightarrow qd\epsilon(\omega) = -\epsilon,
\end{equation}
with $\epsilon$ the dielectric constant of the surrounding background and $\epsilon(\omega)$ that of the polar dielectric. The right-hand side is obtained in the limit of $d\rightarrow 0$. Taking Equation (13) for the dielectric function in bulk, we have that (in the absence of losses) $\epsilon + qd\epsilon_{\infty} + q\frac{NQ^{*2}}{2\epsilon_0 M^*A}\frac{1}{\omega_{TO}^2-\omega^2} = 0$, where we have identified $V = 2Ad$. In the limit of small $d$, i.e., $qd \ll 1$, we recover Equations (16) and (18) with $\Pi = \frac{NQ^{*2}}{\epsilon_0 MA}\frac{1}{\omega^2_{TO}-\omega^2}$. A useful result which follows is that we may also derive another important figure of merit, the $p-$polarized reflectivity of a 2D polar dielectric, which is related to the local density of states of the polar dielectric, which quantifies the strength of coupling to external probes such as a nano-antenna or a quantum emitter. By considering the infinitely thin limit of a polar dielectric with dielectric function $\epsilon(\omega)$ and defining $\epsilon(\omega) = \frac{\Pi}{2d}$, we immediately find that the p-polarized reflectivity is given by
\begin{equation}
r_p = \frac{1}{1+\frac{2\epsilon}{q\Pi}}.
\end{equation}

Before concluding, we note that in general, when considering bulk and 2D versions of a material, the effective parameters $Q^*$ and $M$ will change as a result of changes to the phonon band structure. The losses may also change drastically in the reduction of dimensions. For example, it was found in several experimental works that the damping rate of the TO phonon in hexagonal boron nitride nearly doubles upon going from bulk to two-dimensions \cite{gorbachev2011hunting,tran2016quantum}, which may strongly suppress the ability to image the polariton with near-field probes. This is because due to the small thickness, the dispersion of the phonon polariton is quite flat, and very high wavevector modes, corresponding to the maximum wavevector detectable by a near-field tip, occur at frequencies very close to the TO phonon frequency. Near the TO phonon-frequency however, the ratio $\frac{\text{Im }\Pi}{\text{Re }\Pi}$ is the highest, corresponding to very low quality-factor modes. In addition to these considerations, the values of the parameters may depend on the background material, which may change the phonon band structure.

We provided this thin-film derivation as a simple way to understand the content of Equation (16). In general Equation (17) should be used to calculate $\Pi$ and the corresponding dispersion. We note that a similar situation appears in the treatment of 2D plasmons: namely that a relation like (16) appears, except with $\Pi$ the polarization-polarization response of a 2D electron gas. And furthermore that one can also obtain the same equation by considering the ultra-thin limit of a 3D plasmonic material \cite{jablan2009plasmonics,jablan2013plasmons}.

%\textcolor{blue}{Still looking for more to say about this. This probably requires some brainstorming discussion between us. For now, what I'll do is prepare a plot of the Fourier transform of the green's function of the Maxwell equations for the 2D sheet (p-polarized reflectivity, as was done in Argentene).}
